To be FAIR: Theory Specification Needs an Update

### 2025-03-03 manuscript.Rmd
Fix some typos

Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processeses of theory construction and improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories. Others have used the term "abduction", coined by Peirce to describe "inference to the best explanation". However, Peirce later acknowledged that abduction "is not essentially different from induction" [@peirceCollectedPapersCharles1960, p.4881]. As defining what makes a theory "best" requires auxiliary assumptions, we consider abduction to be a special case of induction. For present purposes, however, the terms are interchangeable.].
<!-- Peirce: CP 8.233 Cross-Ref:†† -->
<!-- 233. This is not essentially different from induction. Only it is most -->
<!-- usually an induction from instances which are not discrete and numerable. I now -->
<!-- call it Qualitative Induction. It is this which I used to confound with the second -->
<!-- line of procedure, or at least not to distinguish it sharply. -->
The present paper restores balance by applying, for the first time,
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing *FAIR theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

The present paper introduces open science methods for theory specification and archival.
Merely publishing theory in a research article does not make it open;
to be open, theory should adhere to established open science standards.
We do so by implementing theories as information artifacts,
archived with apropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
Metadata are "data about the data", they provide information about the nature and content of an information artifact.
Metadata are stored in the repository where the version of record of the FAIR theory is deposited.
FAIR theories are *Findable* via a DOI or by searching the repository they are archived in;
*Accessible* in a machine- and human-readable filetype;
*Interoperable* for specific purposes, for example, within the data analysis environment;
and *Reusable* in the practical and legal sense, so that they may be iteratively improved by the author or by others.
<!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover--> 
<!-- CJ: Since this is in the next paragraph, I think it's redundant. Will merge there -->
Following the original proposal of Lamprecht and colleagues,
we adapt the FAIR principles for theory, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
We reflect on the necessary changes (which are minor),
as well as on the current state and future of FAIR theory in psychology.
The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
and we provide worked examples to encourage their adoption.

@@

FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it increases the fidelity and ease with which they are communicated.
Thus, FAIR theory does not require theories to be formal.
At the same time, formal theories are not automatically FAIR.
The FAIR principles pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility.
They can be applied to theories representated in natural language,
as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
Thus, for example, "grounded theory", derived from qualitative research,
can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
Conversely, a formal theory is not FAIR if it is confined to a journal article without any key words to identify it as a theory paper (lacking Findability), represented merely as a bitmap image (limiting Accessibility and Interoperability), or behind a paywall (limiting Reusability).
FAIR theory is thus consistent with, but does not require, formalization (also see *Accessibility*).

@@

Finally, metadata should be registered or indexed in a searchable registry (F4).
It is important to note that, while may archives are technically searchable (e.g., GitHub, FigShare, the Open Science Framework, institutional repositories),
only few are specifically designed for FAIR-compliant archival.
Zenodo stands out in this respect.
Thus, while using Git for version control and GitHub for collaboration has specific advantages for scientific work [@ramGitCanFacilitate2013],
the version of record should be archived in a FAIR repository like Zenodo.
Using standardized metadata further improves the Findability of theories archived within FAIR repositories.
The DataCite Metadata Schema provides a controlled vocabulary for research output, and the `resource_type: model` matches the description of FAIR theory [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- > An abstract, conceptual, graphical, mathematical or visualization model that represents empirical objects, phenomena, or physical processes. -->
Furthermore, we suggest using the keyword `"fairtheory"` for all resources that constitute or reference (a specific) FAIR theory.
<!-- (separating the words `FAIR` and `theory` by a space or hyphen would lead them to be interpreted as separate tokens in many search engines). -->

@@

There have been notable efforts to improve theories' Findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website [@grayHowMapTheory2017].
Similarly, Van Dongen and colleagues are working on a database of models and formalized theories.
<!-- Is there something to reference? Website/preprint? Otherwise is seems a bit weird to just mention this without any further information. -->
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
Both approaches can be complementary, however.
for example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
Conversely, the database engineering adage "Lots of Copies Keeps Stuff Safe" (LOCKSS) implies that it is fine to archive theories in multiple places,
although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places, which increases the risk of inconsistencies arising.

@@

Interoperability pertains to the property of information artifacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
The common practice of instantiating theory as lengthy prose or schematic drawing falls short of this ideal.
Instead, FAIR theory should, ad minimum,
be instantiated in a human- and machine-readable datatype,
as should all information artifacts created while performing scholarly work [@vanlissaWORCSWorkflowOpen2021].
Depending on the level of formalization of the theory,
different formats may be appropriate,
such as verbal statements in plain text,
mathematical formulae,
and statements expressed in some formal language.
Examples of the latter include pseudo-code,
interpretable computer code,
and Gray's theory maps [@grayHowMapTheory2017].
While a theory represented as a bitmap image is not very Interoperable,
the same image represented in the DOT language [@DOTLanguage2024]
for representing graphs does meet this ideal.

@@

If we compare it to the levels of theory formalization [@guestHowComputationalModeling2021],
it is defined at either the "theory" or "specification" level.
We can increase the level of formalization, and present an "implementation" in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)).
The implementation below describes the model as a directed graph.
Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
and the second half describes their proposed interrelations.
This follows the first two properties of good theory according to Meehl [@meehlAppraisingAmendingTheories1990].
We can save this implementation of the empirical cycle to a text file, say `empirical_cycle.dot`.

@@

Create a new folder and copy the theory file from the previous step into it.
To help meet the Interoperability and Reusability criteria,
add two more files:
A README.md file with instructions for future users of your theory,
and a LICENSE file with the legal conditions for reuse.
For guidance on writing the README file, see [this vignette](https://cjvanlissa.github.io/theorytools/articles/readme.html)
We recommend the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).

>>>>

Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and improvement have been largely overlooked^[We use induction to describe inferences from specific observations to general theories. Others have used the term "abduction", coined by Peirce to describe "inference to the best explanation". However, Peirce later acknowledged that abduction "is not essentially different from induction" [@peirceCollectedPapersCharles1960, p.4881]. As defining what makes a theory "best" requires auxiliary assumptions, we consider abduction to be a special case of induction. For present purposes, however, the terms are interchangeable.].
<!-- Peirce: CP 8.233 Cross-Ref:†† -->
<!-- 233. This is not essentially different from induction. Only it is most -->
<!-- usually an induction from instances which are not discrete and numerable. I now -->
<!-- call it Qualitative Induction. It is this which I used to confound with the second -->
<!-- line of procedure, or at least not to distinguish it sharply. -->
The present paper restores balance by applying, for the first time,
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing *FAIR theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

The present paper introduces open science methods for theory specification and archival.
Merely publishing theory in a research article does not make it open;
to be open, theory should adhere to established open science standards.
We do so by implementing theories as information artifacts,
archived with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
Metadata are "data about the data", they provide information about the nature and content of an information artifact.
Metadata are stored in the repository where the version of record of the FAIR theory is deposited.
FAIR theories are *Findable* via a DOI or by searching the repository they are archived in;
*Accessible* in a machine- and human-readable filetype;
*Interoperable* for specific purposes, for example, within the data analysis environment;
and *Reusable* in the practical and legal sense, so that they may be iteratively improved by the author or by others.
<!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover--> 
<!-- CJ: Since this is in the next paragraph, I think it's redundant. Will merge there -->
Following the original proposal of Lamprecht and colleagues,
we adapt the FAIR principles for theory, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
We reflect on the necessary changes (which are minor),
as well as on the current state and future of FAIR theory in psychology.
The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
and we provide worked examples to encourage their adoption.

@@

FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it increases the fidelity and ease with which they are communicated.
Thus, FAIR theory does not require theories to be formal.
At the same time, formal theories are not automatically FAIR.
The FAIR principles pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility.
They can be applied to theories representated in natural language,
as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
Thus, for example, "grounded theory", derived from qualitative research,
can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
Conversely, a formal theory is not FAIR if it is confined to a journal article without any key words to identify it as a theory paper (lacking Findability), represented merely as a bitmap image (limiting Accessibility and Interoperability), or behind a paywall (limiting Reusability).
FAIR theory is thus consistent with, but does not require, formalization (also see the section on *Accessibility* below).

@@

Finally, metadata should be registered or indexed in a searchable registry (F4).
It is important to note that, while many archives are technically searchable (e.g., GitHub, FigShare, the Open Science Framework, institutional repositories),
only few are specifically designed for FAIR-compliant archival.
Zenodo stands out in this respect.
Thus, while using Git for version control and GitHub for collaboration has specific advantages for scientific work [@ramGitCanFacilitate2013],
the version of record should be archived in a FAIR repository like Zenodo.
Using standardized metadata further improves the Findability of theories archived within FAIR repositories.
The DataCite Metadata Schema provides a controlled vocabulary for research output, and the `resource_type: model` matches the description of FAIR theory [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- > An abstract, conceptual, graphical, mathematical or visualization model that represents empirical objects, phenomena, or physical processes. -->
Furthermore, we suggest using the keyword `"fairtheory"` for all resources that constitute or reference (a specific) FAIR theory.
<!-- (separating the words `FAIR` and `theory` by a space or hyphen would lead them to be interpreted as separate tokens in many search engines). -->

@@

There have been notable efforts to improve theories' Findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website [@grayHowMapTheory2017].
Similarly, Van Dongen and colleagues are working on a database of models and formalized theories.
<!-- Is there something to reference? Website/preprint? Otherwise is seems a bit weird to just mention this without any further information. -->
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
Both approaches can be complementary, however.
For example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
Conversely, the database engineering adage "Lots of Copies Keeps Stuff Safe" (LOCKSS) implies that it is fine to archive theories in multiple places,
although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places, which increases the risk of inconsistencies arising.

@@

Interoperability pertains to the property of information artifacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
The common practice of instantiating theory as lengthy prose or schematic drawing falls short of this ideal.
Instead, FAIR theory should, ad minimum,
be instantiated in a human- and machine-readable datatype,
as should all information artifacts created while performing scholarly work [@vanlissaWORCSWorkflowOpen2021].
Depending on the level of formalization of the theory,
different formats may be appropriate,
such as verbal statements in plain text,
mathematical formulae,
and statements expressed in some formal language.
Examples of the latter include pseudo-code,
interpretable computer code,
and Gray's theory maps [@grayHowMapTheory2017].
While a theory represented as a bitmap image is not very Interoperable,
the same image represented in the DOT language [@DOTLanguage2024]
for representing graphs does meet this ideal (an example of such a DOT representation is given below).

@@

If we compare it to the levels of theory formalization [@guestHowComputationalModeling2021],
it is defined at either the "theory" or "specification" level.
We can increase the level of formalization, and present an "implementation" in the human- and machine-readable DOT language (and thereby fulfill criterion I1 of [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv)).
The implementation below describes the model as a directed graph (see also Figure \@ref(fig:figecs)a).
Note that the code has been organized so that the first half describes an ontology of the entities the theory postulates,
and the second half describes their proposed interrelations.
This follows the first two properties of good theory according to Meehl [@meehlAppraisingAmendingTheories1990].
We can save this implementation of the empirical cycle to a text file, say `empirical_cycle.dot`.

@@

Create a new folder and copy the theory file from the previous step into it.
To help meet the Interoperability and Reusability criteria,
add two more files:
A README.md file with instructions for future users of your theory,
and a LICENSE file with the legal conditions for reuse.
For guidance on writing the README file, see [this vignette](https://cjvanlissa.github.io/theorytools/articles/readme.html).
We recommend the `CC0` license, but other options are available, see [https://choosealicense.com](https://choosealicense.com/non-software/).
